Rapid technological progress in white light-emitting diodes and its sources in innovation and technology spillovers

### 2023-08-13 2_SSL_NE/article/article_main.tex
added edits

There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{Arrow1971}, and economies of scope \cite{national2016power,iea2020perspectives}. The role of these factors at different stages of the innovation life cycle \cite{grubler2012policies}, from research and technology development to demonstration, market formation and diffusion, is an area of active research \cite{Mowery1979} \cite{kavlak2018evaluating} \cite{Ziegler2021}. Innovation is also driven by forces of supply and demand. Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.

The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies,Anadon2016}. Among these drivers, the role of external  knowledge transfer and technology spillovers in research and development of energy technologies is an understudied area  \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context and the research question \cite{Nemet2012}, we follow Stephan et al. \cite{Stephan2021} and consider knowledge to be external - thus a spillover -  if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network outside a technology may be an important factor in understanding \cite{Pichler2020} and shaping \cite{Clark2016,Stephan2021,Sun2021,kolesnikov2022technology} the future evolution of technologies.

Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996, lighting technology has experienced dramatic efficiency improvements.  As shown in Figure \ref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

@@

These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long,stegmaier2021incandescent}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

Despite this impressive history, the sources of LED innovation have not received as much attention from researchers as innovation in supply-side  energy technologies, such as solar photovoltaics  \cite{kavlak2018evaluating} or wind energy \cite{qiu2012price,jennings2020policy}, or in lithium-ion batteries for transportation \cite{Ziegler2021,Stephan2021}. To the best of our knowledge, no study has comprehensively discussed the sources or extent of progress across various metrics of LED cost or performance since the introduction of first commercial white LED products.  Understanding the extent to which individual innovations and knowledge spillovers  contributed to improvements in LED technology, how this effect compares to other sources of improvements such as economies of scale, and how these innovations occurred (i.e., by what mechanisms and actors) will provide valuable lessons both for other demand-side technologies and overall for accelerating clean energy innovation.

To address these questions, in this paper we identify a set of metrics suitable for tracking the historical progress of LED lighting technology. We then trace device efficiency and cost improvements, as well as changes in relevant consumer experience metrics from the time of introduction of the first commercial warm white LEDs in 2003 to 2020, the year with the most recent data available at the time of writing. Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic literature review of scientific literature and industry reports with patent analysis and a series of elite interviews with eminent experts from academia, public research institutions and industry. This approach allows us to identify innovations in white LED technology and examine their origins to discern technology spillovers among them. Data on performance improvements associated with individual innovations identified in patents, scientific publications, industry reports, and interviews is then augmented  by our calculations of device efficiency, including contributions of specific spillovers to the overall LED efficiency (see Supplementary Information for more detail). We further calculate LED manufacturing costs using a bottom-up cost model with process-step resolution that we developed. Finally, we discuss what our observations may mean for our understanding of the innovation process in innovation process in LED technology and for broader policy and industry efforts to accelerate clean energy innovation.

>>>>

There are a range of mechanisms that contribute to improvements in a technology over time, including targeted research and development (R\&D) efforts, economies of scale, learning by doing \cite{WRIGHT_1936, Arrow_1962}, and economies of scope \cite{johansson2012global, national2016power, iea2020perspectives}. The role of these factors at different stages of the innovation life cycle, from research and technology development to demonstration, market formation and diffusion \cite{grubler2012policies}, is an area of active research  \cite{kavlak2018evaluating, Ziegler2021}. Innovation is also driven by forces of supply and demand \cite{Mowery1979}. Various \textit{"technology-push"} drivers reduce the costs of innovating, while \textit{"market-pull"} drivers increase the pay-offs from investing in innovation \cite{anadon2009policy}.

The stages and drivers of the innovation life-cycle for a particular technology play out within a broader innovation system \cite{grubler2012policies, Anadon2016}. Among these drivers, the role of external knowledge and technology spillovers in research and development of energy technologies remains insufficiently studied \cite{Stephan2021}. While the exact definition of spillovers in the literature depends on the context \cite{Nemet2012, kolesnikov2020novel}, we specifically follow the approach of two previous studies \cite{Stephan2021, kolesnikov2020novel} by considering technology spillovers as the application of external knowledge in a technology where knowledge is considered external  if it has been developed for application in other technologies, sectors, or scientific disciplines. There is emerging evidence that understanding spillovers and the knowledge network beyond a particular technology may be an important factor in understanding \cite{Pichler2020, iea2020cleanenergy} and shaping \cite{Clark2016, Stephan2021} the future evolution of technologies.

Among demand-side technologies, the provision of lighting is a particularly important area for climate change mitigation efforts, as it currently   accounts for 15-19\% of global electricity consumption \cite{Zissis2016,doe_electricity}. It is also an area of rapid recent technological change: since the introduction of the first commercial white light-emitting diodes (LEDs) in 1996 \cite{Nakamura2013}, lighting technology has experienced dramatic efficiency improvements.  As shown in  \cref{fgr:history_efficacy}, thanks to the introduction of LED-based solid-state lighting (SSL), the efficacy of lighting sources has increased by three orders of magnitude in just over 20 years, which is significantly faster than the historical progress observed in previous lighting technologies \cite{weinold2021quantifying}. For comparison, the highest performing light-emitting devices today reach efficacies of 220 lm/W \cite{lumistrips2021mid}, while an incandescent light bulb can only reach efficacies of up to 18 lm/W. Moreover, this rapid improvement in efficiency has been accompanied by a similarly impressive decrease in LED manufacturing costs and retail prices. Figure \ref{fgr:cost_lamp_small} shows how LED retail prices have fallen by two orders of magnitude, at an annual price per flux decline of 27.3\% during the 2008-2020 period, in line with previous estimates \cite{Gerke2020}.

@@

These dramatic improvements in lighting technology, supported by the introduction of lighting efficiency regulations phasing out incandescent lightbulbs and targeted policies stimulating LED adoption in many countries, led to the rapid expansion and diffusion of SSL technologies \cite{weinold2020long, stegmaier2021incandescent, Mills2014}. As a result, by 2020, highly efficient LED luminaires were saving an estimated 131 TWh/year in the EU \cite{eu2019impactass} and 442 TWh/year for the US \cite{guidehouse2020adoption}, which is on par with the amount of energy produced annually by all solar photovoltaic installations in these regions. Notably, market adoption of LED lighting is not limited to developed economies \cite{Kamat2020}. For example, durability, low up-front cost and high efficiency of LED light sources have led to their widespread adoption in rural West African communities without access to grid electricity \cite{Bensch2017}. LEDs have also been used in a wide range of applications beyond lighting, such as personal health monitors \cite{Wyatt2020}, watches and smartphones \cite{Bai2017}, potable water treatment \cite{Lui2014}, high-bandwidth wireless data transmission \cite{Haas2016}, and augmented reality eye wear \cite{Lee2016}.

Despite this impressive history, the sources of technological progress in white LEDs have not received as much attention from researchers as innovation  in supply-side energy technologies, such as solar photovoltaics \cite{kavlak2018evaluating, nemet2019solar}, wind energy \cite{qiu2012price, jennings2020policy}, or lithium-ion batteries \cite{Ziegler2021, Stephan2021}. To the best of our knowledge, no study has comprehensively discussed the sources or extent of historical progress across a range of metrics related to LED cost and performance since the introduction of the first commercial white LED products (see Supplementary Note 1 for a brief review of previous literature on this topic). This is consistent with previous observations regarding the marginalization of end-use technologies in the analysis of energy innovation for climate change impact mitigation \cite{Wilson2012, Creutzig2018}. Understanding the extent to which individual innovations and technology spillovers contributed to improvements in white LED technology, how their effect compares to other sources of improvements such as economies of scale, and how these innovations and spillovers occurred (i.e., by what mechanisms and actors) will provide valuable lessons for innovation in other demand-side energy technologies and clean energy innovation in general.

To address these questions, first, we identify three sets of metrics suitable for tracking the historical progress of LED lighting technology: 1) metrics of energy efficiency of LED devices, including the total device efficiency (“lamp efficiency”) and the sub-efficiencies that describe different energy loss channels in a LED device; 2) metrics of lighting quality relevant to consumer experience, including colour rendering index (CRI) and colour temperature (CT); and 3) LED device manufacturing cost. The rationale for the choice of these metrics, along with their detailed descriptions and definitions, are discussed in Supplementary Note 2. We then trace the historical improvements in these metrics, along with associated changes in LED chip architectures, from the time of introduction of the first commercial warm white GaN-based LEDs in 2003 to 2020, the year with the most recent data available at the time of data collection. Given the proprietary nature of knowledge in the SSL industry, we collect corresponding information using a multi-method approach combining a systematic review of scientific literature and industry reports with patent analysis and a series of elite interviews \cite{tansey2009process} with eminent experts from academia, public research institutions and industry (see Methods section and Supplementary Note 3 for details). This approach allows us to identify individual innovations in white LED technology, examine their knowledge origins to discern technology spillovers among them and quantify their contributions to the overall progress in LED efficiency and consumer experience metrics based on our modeling of LED performance and spectral metrics. We further develop a bottom-up LED manufacturing cost model with process-step resolution and use it to analyze the progress in LED manufacturing cost for the “classical” GaN-based LED chip architecture with lateral current spreading shown in Panel A in \cref{fgr:chip_architecture_overview}.
